Parsing Postfix log files

### 2007-05-24 doc/logparser.tex
Documented actions and identifying more aborted mails (arising as a result of restriction_start changes).

    \item [description] A longer, more detailed description of the rule.

    \item [program] The program (smtp, smtpd, qmgr, etc.) whose log lines
        the rule applies to.  This avoids needlessly trying rules which
        won't match the line, or worse, might match unintentionally.

@@

@@

XXX USE THE SAME DOCUMENTATION HERE AS IN Parser.pm

This list contains all the actions available in the algorithm, in the order
in which they tend to occur in practice.

@@

    \item [ignore] Do nothing with this line, merely parse it so that
        unparsed lines can be reported.

    \item [connect] Handle a connection from a remote client.  Complains if
        a connection already exists for this client

    \item [restriction\_start]

    \item [save\_by\_pid]

    \item [clone]

    \item [disconnect]

    \item [save\_by\_queueid]

    \item [qmgr\_chooses\_mail] 
        however qmgr selects every mail for delivery, so most of the time
        this action merely falls through to saving the information from the
        log line (action: save\_by\_queueid; flowchart:~10).

    \item [track]

    \item [commit]

    \item [pickup]

@@

>>>>

    \item [description] Something must have occurred to cause Postfix to
        log each line (e.g. a remote client connecting causes a connection
        line to be logged).  The description field is generally used to
        describe the action causing the log lines this rule matches.

@@

    \item [program] The program (postfix/smtp, postfix/smtpd, postfix/qmgr,
        etc.) whose log lines the rule applies to.  This avoids needlessly
        trying rules which won't match the line, or worse, might match
        unintentionally.

@@

This list contains all the actions available in the algorithm.  Each action
is passed the same arguments:

@@

    \item [\$line] The log line, with the standard syslog(3) fields parsed,
        removed and made available.

    \item [\$rule] The matching rule

    \item [@matches] The fields in the line captured by \$rule's regex.

    \item [IGNORE] IGNORE just returns successfully; it is used when a line
        needs to be parsed for completeness but doesn't either provide any
        useful data or require anything to be done.

    \item [CONNECT] Handle a remote client connecting: create a new state
        table entry (indexed by smtpd pid) and save both the client
        hostname and IP address.

    \item [DISCONNECT] Deal with the remote client disconnecting: enter the
        connection in the database, perform any required cleanup, and
        delete the connection from the state tables.

    \item [SAVE\_BY\_PID] Use the pid from \$line to find the correct
        connection and call \$self->save() with the appropriate arguments -
        see save() in SUBROUTINES for more details. If the connection
        doesn't exist a connection marked faked will be created and a
        warning issued.

    \item [SAVE\_BY\_QUEUEID] Use the queueid from \$rule and @matches to
        find the correct connection and call \$self->save() with the
        appropriate arguments - see save() in SUBROUTINES for more details.
        If the connection doesn't exist a connection marked faked will be
        created and a warning issued.

    \item [COMMIT] Enter the data into the database. Entry may be postponed
        if the mail is a child waiting to be tracked.

    \item [TRACK] Track a mail when it is forwarded to another mail server;
        this happens when a local address is aliased to a remote address.
        TRACK will be called when dealing with the parent mail, and will
        create the child mail if necessary. TRACK checks if the child has
        already been tracked, either with this parent or with another
        parent, and issues appropriate warnings in either case. Tracking
        children is discussed extensively in the paper written about this
        parser; details about obtaining the paper are given in the SEE ALSO
        section.

    \item [REJECTION] Deal with postfix rejecting an SMTP command from the
        remote client: log the rejection with the accepted mail if there is
        one, otherwise log it with the connection.

    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
        picking a mail from the queue to deliver. This action is used for
        both pickup and cleanup due to out of order log lines.

    \item [PICKUP] Pickup is the service which deals with mail submitted
        locally via /usr/sbin/sendmail. This action creates a new state
        table entry and saves data to it, unless out of order logging has
        caused the cleanup line to be logged first. Lines are assumed to be
        out of order if the only program seen thus far is cleanup and there
        is less than five seconds difference between the timestamps of the
        two lines.

    \item [CLONE] Multiple mails may be accepted on a single connection, so
        each time a mail is accepted the connection's state table entry
        must be cloned; if the original data structure was used the second
        and subsequent mails would corrupt the data structure.

    \item [MAIL\_TOO\_LARGE] When the client tries to send a larger message
        than the local server accepts the mail will be discarded.  See
        TIMEOUT for further discussion; the two are handled in exactly the
        same way.

    \item [TIMEOUT] The connection timed out so the mail currently being
        transferred must be discarded. The mail may have been accepted, in
        which case there's a data structure to dispose of, or it may not in
        which case there's none. The gory details can be found in the
        internals documentation.

@@

\subsubsection{Further aborted delivery attempts}

Some mail clients disconnect abruptly if a second or subsequent recipient
is rejected; they may also disconnect after other errors, but such
disconnections are already being handled.  Sadly Postfix doesn't log a
message saying the mail has been discarded, as should be expected by now.
The checks to identify this happening are:

    \item Were there three or more smtpd log lines for the mail?  There
        should be a connection line and a mail accepted line, followed by
        one or more rejection lines.

    \item Is the last smtpd log line a rejection line?

These checks are made during the DISCONNECT action: if all checks are
successful then the mail is assumed to have been discarded when the client
disconnected.  There will be no further entries logged for such mails, so
without identifying and discarding them they accumulate in the state table
and will cause clashes if the queueid is reused.
